---
title: "Exercise 2: Inverse transform sampling"
output:
  html_document:
    toc: false
    number_sections: false
---

```{r, include=FALSE, }
knitr::opts_chunk$set(tidy = TRUE, message = FALSE, cache=TRUE)
```



 1. Program a sampling algorithm to sample from the exponential distribution with parameter $\lambda$ thanks to the inverse transform function (starting from the `R` function `runif`).  
Compare the distribution of your sample to the theoretical target distribution (thanks to the built-in `R` function `dexp`).  
Try out several values for the $\lambda$ parameter of the exponential distribution (e.g. 1, 10, 0.78, ...).

    ```{r Exp, eval = FALSE, echo = TRUE, fig.align = "center", out.width = "67%"}
    generate_exp <- function(n, lambda) {
      u <- ...
      x <- ... u ...
      return(x)
    }

    n_samp <- 1000
    my_samp <- generate_exp(n = n_samp, lambda = 10)
    hist(my_samp, probability = TRUE, n = 25)
    curve(dexp(x, rate = 10), from = 0, to = max(my_samp), col = "red", lty = 2, add = TRUE)
    legend("topright", c("Inverse transform", "R dexp()"), lty = c(1, 2), col = c("black", "red"))
    ```
