---
title: "Exercise 5: Post-mortem analysis of an under-powered randomized trial"
output:
  html_document:
    toc: false
    number_sections: false
---


```{r, include=FALSE, }
knitr::opts_chunk$set(tidy = TRUE, message = FALSE, cache=FALSE)
```

The randomized clinical trial *EOLIA*^[@combes2018Extracorporeal] evaluated a new treatment for severe acute respiratory distress syndrome (severe ARDS) by comparing the mortality rate after 60 days among 249 patients randomized between a control group (receiving conventional treatment, i.e. mechanical ventilation) and a treatment group receiving extracorporeal membrane oxygenation (ECMO) --- the new treatment studied. A frequentist analysis of the data concluded to a Relative Risk of death of $0.76$ in the ECMO group compared to controls (in Intention to Treat), with $CI_{95\%} = [0.55 , 1.04]$ and the associated p-value of $0.09$.

Goligher *et al.* (2019) ^[@goligher2018Extracorporeal] performed a Bayesian re-analysis of these data, further exploring the evidence and how it can be quantified and summarized with a Bayesian approach.


|                                   | Control | ECMO |
|-----------------------------------|---------|------|
| **$n$ observed**                  | 125     | 124  |
| **number of deceased at 60 days** |  57     |  44  |
Table: Observed data from the *EOLIA* trial


 1. Write the Bayesian model used by Goligher *et al.* (2019).
  
    `r if(correction){":::Correction
    ***I) Question of interest:***  
    *Is the Relative Risk of death under ECMO compared to the conventional mechanical treatment less than one ?*  
    <br>
    ***II) Sampling model:***  
    *Let $Z_{control}$ be the number of death in the control group, and $Z_{ecmo}$ the number of death in the ECMO group*
    $$Z_{control} \\sim Binomial(p_c, 125)$$
    $$Z_{ecmo} \\sim Binomial(RR\\times p_c, 124)$$
    <br>
    ***III) Priors:***
    $$p_c\\sim U_{[0,1]}$$
    $$log(RR)\\sim U_{[-35,35]}$$
    <br>
    ***NB:*** One can also define a sampling model at the individual level:  
    *Let $Y_{control_i}$ be a binary variable indicating whether the patient $i$ 
    from the control group died before 60 days, and $Y_{ecmo_i}$ a similar variable for patient from the ecmo group.*
    $$Y_{control_i} \\overset{iid}{\\sim} Bernoulli(p_c)$$
    $$Y_{ecmo_i} \\overset{iid}{\\sim} Bernoulli(RR\\times p_c)$$
    :::
    <br>"}`
  
  
 2. Write the corresponding BUGS model, and save it into a `.txt` file (for instance called `goligherBUGSmodel.txt`)
  
    `r if(correction){":::Correction
    *As we have seen above, there are two equivalent ways of defining the sampling model:  *
      - *either at the population level with a **Binomial** likelihood*,
      - *or at the individual level with a **Bernoulli** likelihood*
    :::
    <br>"}`
    
       ```{BUGS goligher-popmodel-noinfo, eval=FALSE, echo=correction}
       # Population model
       model{
      
         # Sampling model
         zcontrol~dbin(pc, ncontrol)
         zecmo~dbin(RR*pc, necmo)
      
         # Prior
         logRR~dnorm(0, 1/100)
         pc~dunif(0,1) #probability of death in the control group
        
         # Re-parameterizations
         RR <- exp(logRR)
         ARR <- RR*pc
       }
      ```
 
      ```{BUGS goligher-indivmodel-noinfo, eval=FALSE, echo=correction}
      # Individual model
      model{
      
        # Sampling model
        for (i in 1:ncontrol){
          ycontrol[i]~dbern(pc)
        }
        for (i in 1:necmo){
          yecmo[i]~dbern(RR*pc)
        }
      
        # Prior
        logRR~dnorm(0, 1/100)
        pc~dunif(0,1) #probability of death in the control group
        
        # Re-parameterizations
        RR <- exp(logRR)
        ARR <- RR*pc
      }
      ```



 3. First create two binary data vectors `ycontrol` and `yecmo`, that are either `1` or `0`, to encode the observations from the data table above. Then uses the `jags.model()` and `coda.samples()` to replicate the estimation from Goligher *et al.* (2019) (**ProTip:** use the function `window()` to remove the burn-in observation from the output of the `coda.samples` function.)
      ```{r goligher-data-fit, eval = correction, echo = correction, fig.width = 12, fig.height = 8, out.width = "95%"}
        #Individual data
        ycontrol <- c(rep(0, 125-57), rep(1, 57))
        yecmo <- c(rep(0, 124-44), rep(1, 44))
        
        #sampling
        library(rjags)
        goligher_jags_indiv <- jags.model(file = "goligherBUGSmodel_indiv.txt", 
                                          data = list("ycontrol" =  ycontrol, 
                                                      "ncontrol" = length(ycontrol),
                                                      "yecmo" =  yecmo, 
                                                      "necmo" = length(yecmo)
                                                      ), 
                                          n.chains = 3)
        res_goligher_indiv <- coda.samples(model = goligher_jags_indiv, 
                                               variable.names = c('pc', 'RR'), 
                                               n.iter = 40000)
        
        #postprocessing
        res_goligher_burnt_indiv <- window(res_goligher_indiv, start=20001) # remove burn-in for Markov chain convergence
        
        
        
        
        #Population data
        zcontrol <- 57
        zecmo <- 44
        #sampling
        goligher_jags_pop <- jags.model(file = "goligherBUGSmodel_pop.txt", 
                                          data = list("zcontrol" =  zcontrol, 
                                                      "ncontrol" = 125,
                                                      "zecmo" =  zecmo, 
                                                      "necmo" = 124
                                                      ), 
                                          n.chains = 3)
        res_goligher_pop <- coda.samples(model = goligher_jags_pop, 
                                               variable.names = c('pc', 'RR'), 
                                               n.iter = 40000)
        
        #post-processing
        res_goligher_burnt_pop <- window(res_goligher_pop, start=20001) # remove burn-in for Markov chain convergence
        ```
        


  4. Check the convergence, and then comment the estimate results (**ProTip:** look at the effective sample size with the `effectiveSize()` `R` function).
        ```{r goligher-diagnoses, eval = correction, echo = correction}
        effectiveSize(res_goligher_burnt_pop)
        plot(res_goligher_burnt_pop)
        gelman.plot(res_goligher_burnt_pop)
        acfplot(res_goligher_burnt_pop)
        par(mfrow=c(3, 2))
        cumuplot(res_goligher_burnt_pop, ask=FALSE, auto.layout = FALSE)
        par(mfrow=c(1, 1))
        
        summary(res_goligher_burnt_pop)
        summary(res_goligher_burnt_indiv)
        
        
        # shortest 95% Credibility interval:
        HDInterval::hdi(res_goligher_burnt_pop) 
        
        # posterior porbability of RR <1:
        mean(c(sapply(res_goligher_burnt_pop, "[", , 1))<1)
        ```


  5. Change to a more informative *prior* using a Gaussian distribution for the log(RR), centered on log(0.78) and with a standard deviation of 0.15 in the log(RR) scale (i.e. a precision of $\approx 45$). Comment the results. Try out other *prior* distributions.
        ```{BUGS goligher-info, eval=FALSE, echo=correction}
         logRR~dnorm(log(0.78), 45)
        ```